Refinements for Open Automata

### 2023-09-01 2021-Refinement/paper/refinementarticleV2.tex
cont. corrections + integration rapport

Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one, simulation relations formalise the desired relation between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of simulation between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

A  bisimulation relation  relates pair of states and ensures that any behaviour of one automaton can be performed by the other one while staying in relation. We informally explain here the symbolic nature of the bisimulation for OA and the related complexity of its definition. The presence of holes in fact raises no strong difficulty but the variables must be handled carefully.
Consider the two following simple OA:

@@

\begin{definition}[Hole-equal simulation]\label{def:HoleEqualSim}
Consider two OA with identical set of holes:  \(A_1=\OAg[1]\) and \(A_2=\OAg[2]\) where  \(J_1 = J_2\), the relation on configurations \(\mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $A_1$ to $A_2$ if the following conditions hold : 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
 \bigsymb{\forall} {t_1} = \openrule
         {
           \beta_{1 j}^{j\in J}, g_1,\psi_1}
         {s_1 \OTarrow {\alpha_1} s_1'} \in \fOT{s_1} .\,~~
\bigsymb{\bigexists} 
\mpar{t_{2x} \!=\! \openrule
         {
           \beta_{2 x j}^{j\in J}, g_{2 x},\psi_{2 x}}
         {s_2 \OTarrow {\alpha_{2 x}} s_{2 x}'}
}^{x \in X}\in \fOT{s_2}.\,~~\\ 
%\everymath{\displaystyle}\begin{array}{l}
V_1 \uplus V_2 \uplus  \fvars{t_1} \uplus \bigcup_x\fvars{t_{2x}}  \vdash \hspace{7cm}
\\
 \mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies
 \operatorname*{\bigsymb{\bigvee}}_{x \in X}
{\everymath{\displaystyle}
\mpar{\begin{array}{l}
			\alpha_{2x} = \alpha_{1} \wedge \bigwedge_\subbox{j \in J} \beta_{2xj} = \beta_{1j} \nwedge\\[12pt]
			 g_{2x} \wedge\mathcal{R}\mpar{ s'_{1},s'_{2x}}\psubst{\psi_{2x} \uplus \psi_{1}}
		\end{array}} }
%\end{array} 
\end{multline*}
\item [(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \!\vdash\! \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
If there is a hole-equal simulation from $A_1$ to $A_2$, then we say that  $A_2$ simulates $A_1$; we  denote it $A_2\leq A_1$.
\end{definition}

@@

We now  state the properties of our simulation, their proofs  can be found in the appendices. We express these properties  in terms of hole-tracking simulation because, thanks to Property~\ref{lem-sim-sim}   all the properties  of hole-tracking simulation are also valid  for hole-equal simulation.
The first crucial theorem of  simulation  is that it is  a  preorder on the set of OA. This latter enables stepwise refinement.

@@

\begin{proofsketch}
% TODO: I don't understand this definition, why v = v on vars(s1), what about s2
% ANS: was copy paste from the report, polished now
The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\). This is shown  by considering the relation $\mathcal{R}$ such that $\mathcal{R}\mpar{s_1, s_2} \triangleq s_1=s_2\land \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$   we can prove the conditions for Definition~\ref{Def:OA-Refinement}.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
Appendix~\ref{sec:proof-transitivity} presents the proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a simulation. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a  formula that defines the simulation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
  \[\mathcal{R}_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{\mathcal{R}_{12}(s_1,s_2) \land \mathcal{R}_{23}(s_2,s_3) } \]
We then prove that this relation  is a simulation, according to Definition~\ref{Def:OA-Refinement}.
\end{proofsketch}

@@

\newpage

\appendix

\section{Proof of Transitivity for Refinement (Theorem~\ref{thm:preoder})}\label{sec:proof-transitivity}
\begin{quote}
If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
\end{quote}

\proof 
If $\wrel{A_1}{A_2}{H}$ then there is $\mathcal{R}_{12}$ a relation between states
of $A_1$ and of $A_2$;  If $\wrel{A_2}{A_3}{H'}$ then there is $ˇ_{23}$ a relation between states of $A_2$ and of $A_3$. We build a relation between
 states of $A_1$ and of $A_3$ as follows:  for each pair of states $s_1$, $s_3$, for each state $s_2$ such that $\mathcal{R}_{12}$ relates $s_1$ and $s_2$, and $\mathcal{R}_{23}$ relates $s_2$ and $s_3$.
Let $\mathcal{R}_{13}$ be the relation:\\
  \[\mathcal{R}_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{\mathcal{R}_{12}(s_1,s_2) \land \mathcal{R}_{23}(s_2,s_3) } \]

We will show that $\wrel{A_1}{A_3}{H\cap H'}$ by exhibiting  $\mathcal{R}_{13}$ as a hole-tracking simulation of $A_1$ by  $A_3$.

We have to prove that the relation $\mathcal{R}_{13}$ satisfies the three conditions of the definition of a refinement of OA.
\begin{enumerate}
\item Firstly, we have to $\mathcal{R}_{13}$ satisfies initial configurations:
\[\sigma_{01} \uplus \sigma_{03} \vdash \mathcal{R}_{13} \mpar{s_{01}, s_{03}}\]
By knowing that substitutions only have an effect on the variables of the OA they belong to, they also produce terms containing only variables of the OA they belong to. We have:
\begin{align*}
\mpar{\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}_{12}\mpar{s_{01}, s_{02}}} \wedge \mpar
{\sigma_{02} \uplus \sigma_{03} \vdash \mathcal{R}_{23}\mpar{s_{02}, s_{03}}}&\implies\\
\mathcal{R}_{12}\mpar{s_{01},s_{02}}\psubst{\sigma_{01} \uplus \sigma_{02}} \wedge \mathcal{R}_{23}\mpar{s_{02}, s_{03}}\psubst{\sigma_{02} \uplus \sigma_{03}}&\implies\\
\mathcal{R}_{12}\mpar{s_{01},s_{02}}\psubst{\sigma_{01} \uplus \sigma_{02} \uplus \sigma_{03}} \wedge \mathcal{R}_{23}\mpar{s_{02}, s_{03}}\psubst{\sigma_{01} \uplus \sigma_{02} \uplus \sigma_{03}}&\implies\\ 
\underbrace{\mathcal{R}_{12}\mpar{s_{01},s_{02}}\wedge \mathcal{R}_{23}\mpar{s_{02}, s_{03}}}_{\implies \mathcal{R}_{13} \mpar{s_{01}, s_{03}}}\psubst{\sigma_{01} \uplus \sigma_{02} \uplus \sigma_{03}} %&\implies\\
 \end{align*}
Since $\sigma_{02}$ has no effect on variables of $s_{01}$ and $s_{03}$ thus we get the expected result.

\item 
 Secondly, we need to prove that for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{3}$: 
\begin{multline*}
V_1 \uplus  V_3 \uplus \fvars{t_1}
\vdash\\\hspace{2em}
\mpar{R_{13}\mpar{s_1, s_3} \wedge g_1 \implies 
\operatorname*{\bigsymb{\bigvee}}_{z \in Z}
\mpar{\everymath{\displaystyle}
\begin{array}{l}
\alpha_1=\alpha_{3z}\wedge \bigwedge_\subbox{j \in J'_{3z} \cap (H\cap H')} \beta_{1j}=\beta_{3jz} \wedge \\g_{3z}\wedge 
{R_{12}\mpar{s'_1,s'_{3z}}}\psubst{\psi_1  \uplus \psi_{3z}}
\end{array}}}
\end{multline*}	
The proof of this simulation step follows the same principles as the one given in \cite{AMHEEMA:2023} for the proof of transitivity of bisimulation\footnote{In particular, the bisimulation relation exhibited for proving transitivity is the same.}.  It relies on  the fact that one open transition of $A_1$ is simulated by a
family of open transitions of $A_2$, and one open transition of $A_2$ is simulated by a family of open transitions of $A_3$. Transitivity leads to a doubly indexed family of simulating open transition.

\item Lastly, we have to prove the satisfaction of the deadlock reduction condition. 
 To build $\mathcal{R}_{13}$ we need to rely on the disjunction of all possible paths to relate $s_1$ and $s_3$, which leads to \[{\displaystyle \mathcal{R}_{13}(s_1,s_3)=\bigvee_{{p\in P}}\mpar{\mathcal{R}_{12}\mpar{s_1,s_{2p}} \wedge \mathcal{R}_{23}\mpar{s_{2p},s_3}}}\]

Consider any $\mpar{s_{1}, s_{3}} \in \mathcal{R}_{13}$  there is a set of states $(s_{2p})^{p\in P}$ of $A_2$ relating  $s_{1}$ and $s_{3}$.

First, according to the relation between $A_1$ and $A_2$ OA,  for all $\mpar{s_1, s_{2p}} \in S_1\times S_2$  we have:
\begin{multline*}
%\forall \mpar{s_1, s_{2p}} \in S_1 \times S_2, 
V_1 \uplus V_2  \vdash \mathcal{R}_{12}\mpar{s_1, s_{2p}} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}  \lor¢ \lnot( \bigvee_\subbox{t_{2p} \in \fOT{s_{2p}}} \fguard{t_{2p}}  )
\end{multline*}
Second, according to the relation between $A_2$ and $A_3$ OA, for all $\mpar{s_{2p},s_3} \in S_2\times S_3$  we have:
\begin{multline*}
 V_2 \uplus V_3  \vdash \mathcal{R}_{23}\mpar{s_{2p},s_3} \implies  \bigvee_\subbox{t_1 \in \fOT{s_{2p}}} \fguard{t_{2p}}  \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
With the conjunction of both, we get:

\begin{multline*}
 V_1 \uplus V_2 \uplus V_3  \vdash  \mathcal{R}_{12}\mpar{s_1, s_{2p}} \wedge \mathcal{R}_{23}\mpar{s_{2p},s_3} \implies \\
  \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} 
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}

This is valid for all $s_{2p} \in (s_{2p})^{p\in P}$, then we have:
\begin{multline*}
 V_1 \uplus V_2 \uplus V_3  \vdash \underbrace{\bigvee_{{p\in P}}\mpar{R_{12}\mpar{s_1, s_{2p}} \wedge R_{23}\mpar{s_{2p},s_3}}}_{R_{13}(s_1,s_3)} \implies \\ ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
By removing the $A_2$ variables that have not effect on $A_2$ et $A_3$,
we get the desired result:
\begin{multline*}
 V_1 \uplus  V_3  \vdash \mathcal{R}_{13}(s_1,s_3) \implies  ( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} )
 \lor¢ \lnot( \bigvee_\subbox{t_3 \in \fOT{s_3}} \fguard{t_3} )
\end{multline*}
\qed
\end{enumerate}
\section{Proof of Context Refinement (Theorem~\ref{thm:ContextRefinement})}

\begin{quote}
Suppose that $\wrel{A_1}{A_2}{H}$, \(k \in H\) and that \(A_1\subst{A_3}{k}\) is non-blocking. We have: 
\[\wrel{A_1\subst{A_3}{k}}{A_2\subst{A_3}{k}}{J_3 \uplus H \setminus \mbrc{k}}\]
\end{quote}
\smallskip
Pictorially, the theorem states the following result:

\begin{center}
\includegraphics[scale=0.7]{Figures/Thm2intuition}
\end{center}

\proof
Let us denote by $A_{13}$ (resp. $A_{23}$) the OA resulting from $A_1\subst{A_3}{k}$ (resp. $A_2\subst{A_3}{k}$),  to prove the theorem it is sufficient to prove that there exists a relation between states of the two OA that satisfies the conditions of  Definition~\ref{Def:OA-Refinement}.

We denote $A_1=  \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}$ and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\) and $A_3 = \OA{S_3}{s_{03}}{J_3}{V_3}{\sigma_{03}}{T_3}$. The proof requires  to rename the variables of one instance of the two $A_3$ automata to avoid clashes in variable names (this is required by the definition of refinement). In practice we will use superscripts ${}^1$ and ${}^2$ to distinguish elements of the two instances of $A_3$.

Let $\mathcal{R}$ be the refinement relation relating states of $A_1$ and $A_2$. 
Let us denote with $t^1$ and $t^2$  the elements of $A_1$ and $A_2$ respectively.
Consider any two states $s_{13} = \mpar{s_1,s^1_3}$ and $s_{23} = \mpar{s_2,s^2_3}$ ($s^1_3$ and $s^2_3$  are the same with renaming). We define a relation $\mathcal{R}'$ relating states of $s_{13}$ and $s_{23}$ as follows:
\[ \mathcal{R}' \mpar{s_{13}, s_{23}} = \mathcal{R}\mpar{s_{1}, s_{2}} \land \reach{A_{13}}(s_{13})  %\wedge \reach{A_{13}}\mpar{s_{13}}
 \wedge \bigwedge_\subbox{v_3\in V_3}
 v^1_3=v^2_3\land s^1_3 = s^2_3\]

%\TODO{[Ludo] I believe reachability is not useful here -- to check}

We want to prove that $\mpar{\mathcal{R}', H\uplus J_3 \setminus\{k\}}$ is a hole-tracking simulation of $A_{13}$ and $A_{23}$. In the following we denote $H'=H\cup J_3 \setminus\{k\}$.
\begin{enumerate}
\item First, we have to prove the relation for initial states:
\[\sigma_{013} \uplus \sigma_{023} \vdash \mathcal{R}'\mpar{s_{013}, s_{023}}\]
with $\sigma_{013} = \sigma_{01} \uplus \sigma_{03}^1$, $\sigma_{023} = \sigma_{02} \uplus \sigma_{03}^2$, $s_{013}=(s_{01},s_{03}^1)$, and $s_{023}=(s_{02},s_{02}^2)$.\\
By using the fact that $\mathcal{R}$ relates initial configurations of  $A_1$ and $A_2$, we have:
$\mpar{\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}}$.  
%and based on the  fact that initial states are reachable $\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}$ we have 
%\[ \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge 
%\mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}\]

Considering that initial valuations $\sigma_{03}^1$ and $\sigma_{03}^2$ associate the same values to the ``same'' variables modulo renaming, so the following holds:\\ $\left(\,\,{\displaystyle \bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\right)\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }$.

Additionally, because  the domains of the substitution function are disjoint, the substitution function has an effect only on the related elements,  we get:
\begin{align*}
& \mpar{\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}} 
%\wedge \mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}
\\
\implies & \mathcal{R}\mpar{s_{01}, s_{02}}\psubst{ \sigma_{01} \uplus \sigma_{02}}  \land \mpar{\,\,\bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }\\
\implies & \mpar{\mathcal{R}\mpar{s_{01}, s_{02}} \wedge  \bigwedge_\subbox{v_3\in V_3}  v^1_3=v^2_3\wedge s^1_3 = s^2_3} \psubst{ \sigma_{01} \uplus \sigma_{02} \uplus \sigma^1_{03}\uplus \sigma^2_{03}\uplus \sigma_{013} }\\
\implies & \sigma_{013} \uplus \sigma_{023} \vdash \mathcal{R}'\mpar{s_{013}, s_{023}}
\end{align*}
The last step comes from the additional fact that  $\reach{A_{13}}(s_{013})$.

\item Second, we need to prove for any open transition $t_{13}$ in $T_{13}$ originating from $s_{13}$:
\begin{align*}
		&  \OTx{13}{}{13}{13} \in \fOT{s_{13}}
\end{align*}		
there exists an indexed family $t_{23x}$ of OTs originating from $s_{23}$ that simulate it: 
\begin{align*}
		&  \mpar{\OTx{23}{x}{23x}{23x} \in \fOT{s_{23}}}^{x \in X}
\end{align*}	
\begin{align*}		
		& \text{such that } \mpar{\forall x \in X, J'_{23x} \cap H' = J'_{13} \cap H'} \text{ and }\\
		&  V_{13} \uplus V_{23} \uplus \fvars{t_{13}} \vdash\\
		& \mathcal{R}'\mpar{s_{13}, s_{23}} \wedge g_{13} \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}\begin{array}{l}
			\alpha_{13} = \alpha_{23x} \wedge \bigwedge_\subbox{j \in J'_{23x} \cap H'} \beta_{13j} = \beta_{23xj} \wedge\\[12pt]
			 g_{23x} \wedge \mathcal{R}'\mpar{s'_{13}, s'_{23x}}\psubst{\psi_{13} \uplus \psi_{23x}}
		\end{array}} 
	\end{align*}

Recall that by definition of composition and OA refinement we have: 
\begin{align*}
&V_{13}= V_1\uplus V^1_3 \text{ and }
V_{23}= V_2\uplus V^2_3\\
& H' \subseteq J_3\uplus \mpar{J_1 \cap J_2} \setminus \{k\} = \mpar{J_3\uplus J_1 \setminus \{k\} } \cap \mpar{J_3\uplus J_2 \setminus \{k\}} \\
\end{align*}
First of all, we have by hypothesis $\wrel{A_1}{A_2}{H}$, then for any open transition $t_1$ in $T_1$  originating from $s_1$:
\begin{align*}
\OTx{1}{}{1}{1} \in \fOT{s_1}
\end{align*}
there exists an indexed family of OTs originating from $s_{2}$: 
\begin{align*}
\mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_{2}}}^{x \in X} 
\end{align*}
such that $\forall x \in X, J'_{2x} \cap H = J'_1 \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_1 \uplus V_2 \uplus \fvars{t_1} \vdash\\ \mathcal{R}\mpar{s_1, s_{2}} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \nwedge\\
			 g_{2x} \wedge \mathcal{R}\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}  \qquad \quad (*)
\end{multline*}	
\smallskip

Consider any transition $t_{13}$ in $A_{13}$.  Based on  the definition of composition $t_{13}$ can be obtained from two different cases, we will consider the two cases separately.

\paragraph{First case: Both automata perform a transition.}
The transition $t_{13}$ is obtained by the composition of  transitions $t_1=\OTx{1}{}{1}{1}\in \fOT{s_1}$ and  \[t_3^1=
\OT{ s^1_3}{{s^1_3}'}{\alpha_3^1}{   {(\beta^1_{3j})^{j \in J^{1 \prime}_3} }}{ g^1_3 }{ \psi^1_3}\in \fOT{s^1_3} \qquad \text{ when } k \in J'_{1} 
\] The result is:
\[
t_{13} = \OT{\mpar{s_1, s^1_3}}{\mpar{s'_1, {s^1_3}'}}{\alpha_1}{\beta_{1j}^{j \in J'_1 \setminus \{k\}} \uplus  {(\beta^1_{3j})^{j \in J^{1 \prime}_3} }}{g_1 \wedge g^1_3 \wedge \alpha^1_3 = \beta_{1k}}{\psi_1 \uplus \psi^1_3} \qquad \text{ where } k \in J'_{1} 
\]
We then obtain a family of OTs by the simulation of $A_1$ by $A_2$ (as stated above).
By hypothesis we have $k \in H$, so in the case where $k \in J'_{1}$, we deduce that $k \in J'_{2x}$ we can  then build  a family of OTs $t_{23x}^{x\in X}$ with the same transitions of $A_3$ (up to renaming) as those used to build $t_{13}$.
\[
t_{23x} = \left(\OT{\mpar{s_2, s^2_3}}{\mpar{s'_{2x}, {s^2_3}'}}{\alpha_{2x}}{\beta_{2xj}^{j \in J'_{2x}  \setminus \{k\}} \uplus   {(\beta^2_{3j})^{j \in J^{2 \prime}_3} }}{g_{2x} \wedge g^2_3 \wedge \alpha^2_3 = \beta_{2xk}}{\psi_{2x} \uplus \psi^2_3}
\right)^{x\in X}\]
Recall that in this case  $k \in J'_{1}$, so $\forall x \in X$ we have
\begin{align*}
  J'_{23x} \cap H'  
&=((J'_{2x}  \setminus \mbrc{k})\uplus J^{2 \prime}_{3} ) \cap (H\uplus J_3 \setminus\{k\} )\\
&=((J'_{2x}\cap (H\uplus J_3))\uplus (J^{2 \prime}_{3}\cap (H\uplus J_3)))\setminus \mbrc{k}\\
&=((J'_{2x}\cap H)\uplus (J^{2 \prime}_{3}\cap  J_3))\setminus \mbrc{k}\\
& \qquad\qquad\qquad\qquad \text{ since } J_3\cap J'_{2x} = \emptyset \text{ and } H\cap J^{2 \prime}_{3} = \emptyset\\
& =((J'_{2x}\cap H)\uplus J^{2 \prime}_{3})\setminus \mbrc{k}\text{ since } J^{2 \prime}_{3}\subseteq J_3\\
& =((J'_1\cap H)\uplus J^{1 \prime}_{3})\setminus \mbrc{k}\quad \text{ since } J^{1 \prime}_{3}=J^{2 \prime}_{3} \text{ and } J'_1 \cap H = J'_{2x} \cap H\\
&= ((J'_1 \cap H) \uplus (J^{1 \prime}_{3} \cap J_3)) \setminus \mbrc{k}\text{ since } J^{1 \prime}_{3}\subseteq J_3\\ 
& = (J'_1\cap (J_3 \uplus H)) \uplus ((J^{1 \prime}_{3} \cap (J_3 \uplus H)) \setminus \mbrc{k} \\
& \qquad\qquad\qquad\qquad \text{ since } J_3\cap J'_{1} = \emptyset \text{ and } H\cap J^{1 \prime}_{3} = \emptyset\\
		& = \mpar{(J'_1  \uplus J^{1 \prime}_{3})\setminus \mbrc{k}} \cap \mpar{(J_{3} \uplus H) \setminus \mbrc{k}}\\
		& = J'_{13} \cap H'
\end{align*}

>>>>

Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one, simulation relations formalise the desired relationship between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of simulation between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

A  bisimulation relation  relates pair of states and ensures that any behaviour of one automaton can be performed by the other one while staying in relation. We informally explain here the symbolic nature of the bisimulation for OA and the related complexity of its definition.
\RAB{The notion of symbolic bisimulation, as it was introduced in \cite{HENNESSY:1995}, is aimed at computing bisimulation of value-passing processes. The class of processes manipulating and exchanging with the environment data, where the values are from possibly infinite domain.}
 The presence of holes in fact raises no strong difficulty but the variables must be handled carefully.
Consider the two following simple OA:

@@

\begin{definition}[Hole-equal simulation]\label{def:HoleEqualSim}
Consider two OA with identical set of holes:
\(A_1 = \OA{S_1}{s_{01}}{V_1}{\sigma_{01}}{J}{T_1}\) and \(A_2 = \OA{S_2}{s_{02}}{V_2}{\sigma_{02}}{J}{T_2}\), the relation on configurations \(\mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $A_1$ to $A_2$ if the following conditions hold : 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
 \bigsymb{\forall} {t_1} = \openrule
         {
           \beta_{1 j}^{j\in J'_{1}}, g_1,\psi_1}
         {s_1 \OTarrow {\alpha_1} s_1'} \in \fOT{s_1} .\,~~
\bigsymb{\bigexists} 
\mpar{t_{2x} \!=\! \openrule
         {
           \beta_{2 x j}^{j\in J'_{2x}}, g_{2 x},\psi_{2 x}}
         {s_2 \OTarrow {\alpha_{2 x}} s_{2 x}'}
}^{x \in X}\in \fOT{s_2}.\,~~\\ 
%\everymath{\displaystyle}\begin{array}{l}
V_1 \uplus V_2 \uplus  \fvars{t_1} \uplus \bigcup_x\fvars{t_{2x}}  \vdash \hspace{7cm}
\\
 \mathcal{R}\mpar{s_1, s_2} \wedge g_1 \implies
 \operatorname*{\bigsymb{\bigvee}}_{x \in X}
{\everymath{\displaystyle}
\mpar{\begin{array}{l}
			\alpha_{2x} = \alpha_{1} \wedge \bigwedge_\subbox{j \in J} \beta_{2xj} = \beta_{1j} \nwedge\\[12pt]
			 g_{2x} \wedge\mathcal{R}\mpar{ s'_{1},s'_{2x}}\psubst{\psi_{2x} \uplus \psi_{1}}
		\end{array}} }
%\end{array} 
\end{multline*}
\item [(3)] Deadlock reduction:
\[
\forall \mpar{s_1, s_2} \!\in\! S_1 \!\times\! S_2, V_1 \uplus V_2  \!\vdash\! \mathcal{R}\mpar{s_1, s_2} \implies( \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} ) \lor¢ \lnot( \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2}  )
\]
If there is a hole-equal simulation from $A_1$ to $A_2$, then we say that  $A_2$ simulates $A_1$; we  denote it $A_2\leq A_1$.
\end{definition}

@@

We now  state the properties of our simulation, their formal proofs  can be found in  the extended version of this paper \cite{henrio:hal-04193421}. We express these properties  in terms of hole-tracking simulation because, thanks to Property~\ref{lem-sim-sim}   all the properties  of hole-tracking simulation are also valid  for hole-equal simulation.
The first crucial theorem of  simulation  is that it is  a  preorder on the set of OA. This latter enables stepwise refinement.

@@

\begin{proofsketch}
% TODO: I don't understand this definition, why v = v on vars(s1), what about s2
% ANS: was copy paste from the report, polished now

The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\). This is shown  by considering the relation $\mathcal{R}$ such that $\mathcal{R}\mpar{s_1, s_2} \triangleq s_1=s_2\land \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$   we can prove the conditions for Definition~\ref{Def:OA-Refinement}.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
In \cite{henrio:hal-04193421}, we give proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a simulation. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a  formula that defines the simulation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
  \[\mathcal{R}_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{\mathcal{R}_{12}(s_1,s_2) \land \mathcal{R}_{23}(s_2,s_3) } \]
We then prove that this relation  is a simulation, according to Definition~\ref{Def:OA-Refinement}.
\end{proofsketch}

@@
